---
title: "Generating Study Diagnostics"
author: "Martijn J. Schuemie, Patrick Ryan, and Jamie Weaver"
date: "`r Sys.Date()`"
output:
  pdf_document:
    number_sections: yes
    toc: yes
  html_document:
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{Generating Study Diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
options(width = 200)
library(StudyDiagnostics)
knitr::opts_chunk$set(
  cache = FALSE,
  comment = "#>",
  error = FALSE,
  tidy = FALSE)
```
# Introduction

# Cohort Instantiation

Most of the diagnostics included in this package require the cohorts designed in ATLAS to be instantiated in a cohort table. There are several ways in which we can instantiate cohorts:

1. Using ATLAS to instantiate the cohorts. 
2. Using the StudyDiagnostics package to instantiate the cohorts.
3. Using the insertCohortDefinitionSetInPackage function in the ROhdsiWebApi package.

Instantiated cohorts are stored in a cohort table with the fields defined in the Common Data Model. note that this table need not be called 'cohort'. If we are interested in statistics on how many subject meet the various inclusion rules, we also need additional tables to store these results.

## Using ATLAS

Contact your database administrator to identify the database schema (`cohortDatabaseSchema`) and table where the cohorts are instantiated (`cohortTable`). 

For inclusion rule statistics you also need to identify the results database schema (`resultsDatabaseSchema`), which is usually not the same as the cohort database schema. In this database schema there should be the following four tables:

1. `cohortInclusionTable`, e.g. `cohortInclusionTable <- "cohort_inclusion"`
2. `cohortInclusionResultTable`, e.g. `cohortInclusionResultTable <- "cohort_inclusion_result"`
3. `cohortInclusionStatsTable`, e.g. `cohortInclusionStatsTable <- "cohort_inclusion_stats"`
4. `cohortSummaryStatsTable`, e.g. `cohortSummaryStatsTable <- "cohort_summary_stats"`

## Using StudyDiagnostics

```{r eval=FALSE}
createCohortTable(connectionDetails = connectionDetails,
                  cohortDatabaseSchema = cohortDatabaseSchema,
                  cohortTable = cohortTable,
                  createInclusionStatsTables = FALSE)
```

```{r eval=FALSE}
instantiateCohort(connectionDetails = connectionDetails,
                  cdmDatabaseSchema = cdmDatabaseSchema,
                  oracleTempSchema = oracleTempSchema,
                  cohortDatabaseSchema = cohortDatabaseSchema,
                  cohortTable = cohortTable,
                  baseUrl = baseUrl,
                  cohortId = cohortId)
```

## Using insertCohortDefinitionSetInPackage

