---
title: "How to use an external concept counts table"
author: "CE Barboza"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use External Concept Counts Table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
library(CohortDiagnostics)
```

Run the function `createConceptCountsTable()` to save the `concept_counts` in the working schema. 
This is useful since the user doesn't have create that table every time `executeDiagnostics()` is called, saving a significant amount of time when running CohortDiagnostics multiple times on the same vocabulary version. 

```{r eval=FALSE, message=FALSE, warning=FALSE}
CohortDiagnostics::createConceptCountsTable(connectionDetails = connectionDetails,
                                            connection = connection,
                                            cdmDatabaseSchema = cdmDatabaseSchema,
                                            tempEmulationSchema = tempEmulationSchema,
                                            conceptCountsDatabaseSchema = cohortDatabaseSchema,
                                            conceptCountsTable = "concept_counts",
                                            conceptCountsTableIsTemp = FALSE,
                                            removeCurrentTable = TRUE)
```

After creating the table, the user must set the parameter `useExternalConceptCountsTable` as *TRUE* when running `executeDiagnostics()` and specify the name of the table with the parameter `conceptCountsTable`. If `useExternalConceptCountsTable` is set as *FALSE* (the default option), CohortDiagnostics will create a new `concept_count` as usual. 

```{r eval=FALSE, message=FALSE, warning=FALSE}
CohortDiagnostics::executeDiagnostics(cohortDefinitionSet = cohortDefinitionSetRow,
                                      connectionDetails = connectionDetails,
                                      cohortTable = cohortTable,
                                      cohortDatabaseSchema = cohortDatabaseSchema,
                                      cdmDatabaseSchema = cdmDatabaseSchema,
                                      conceptCountsTable = "concept_counts",
                                      exportFolder = exportFolder,
                                      databaseId = databaseId,
                                      minCellCount = minCellCount,
                                      useExternalConceptCountsTable = TRUE)
```